Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-08-26 WSSSPE3/report/JORS-reviews-responses.txt

We thank the reviewers for their insightful comments, which follow, along
with our responses and the changes we've made.

Reviewer A:

Comments to the authors: 
	The paper summarizes the agenda, discussion, and outputs of the WSSSPE3.0
workshop on scientific software sustainability. The paper provides good
insight into the status of that discourse, very relevant to the JoRS venue.
Each section provides useful literature background and identifies people and
groups that readers can follow up with if they are interested in particular
topics. The inclusion of the longer appendices is very useful.  I have two
comments that would improve the article, neither of which I think are
crucial to publication but would be useful.

The paper lays out it's structure at the start, but each section could end
with additional signposts ("having done x, and before doing y, we will
review z"). I think those help the reader know where they are in a long
document.

RESPONSE:

Time has passed since the submission was prepared. I am conscious that my
late review played a role here. However it does create an opportunity to
check in with each of the groups and see if any progress has been made. For
example, it seems as if the effort to create the "best practices" paper has
flagged (at least nothing has been done since November 2015 at the gdoc
linked from the paper. Perhaps ask the conveners/contacts of each working
group if they have any updates?

RESPONSE:

Just a few typos I spotted (very few overall):

p 27 "refunded" should be "re-funded" or "funded again"

RESPONSE: We've changed the word to re-funded.

p2 "Overall, the report discusses the organization work done before the
workshop (§2); the keynote (§3); a series of lightning talks (§4)." needs
some work. Perhaps "Overall, the report discusses the organizational work
done before the workshop (§2); the keynote (§3) and summarizes the
lightning talks presented at the meeting (§4)."

RESPONSE: we've changed this text to
Overall, the report discusses the organization work done before the
workshop (§2), the keynote (§3), and the lightning talks presented
at the meeting (§4).

------------------------------------------------------

------------------------------------------------------
Reviewer B:

Comments to the authors: 
	General Comments:

The authors have provided a dense, lengthy but complete record of the
WSSSPE3 conference.  It follows on the WSSSPE2 report published in JORS, and
is balanced, trustworthy, and valuable. The authors’ limited
editorializing is simultaneously welcome in a historical report and renders
the lengthy document difficult to read or interpret since it lacks a strong
narrative form. To put that a bit more harshly, the manuscript feels like a
network graph with a goal of encompassing all the historical facts around
WSSSPE3. I think this stems from having a dual goal of tracing accurately
what happened at this conference and trying to encode action items and
practical outcomes into an authoritative document about what WSSSPE as an
entity is trying to become. I would recommend that this aspect of the
manuscript could be improved. Most of my concern is centered on Section 5.

Section 5 “Working Groups” beings without any narrative explaining where
these working groups came from and how they might have been culled from a
list of potential/actual working groups. The 10 working groups do not
correlate precisely with the “18 potential topics” on page 3 nor the 6
categories superimposed on the 18 topics. I suppose the working groups could
have been drawn from them though.  So this sets an unanswered question: were
these working groups decided upon by WSSSPE3 conference members or by a
WSSSPE body before the event? (Aside: does a WSSSPE entity exist today? Did
it exist before WSSSPE1? After WSSSPE2?)

RESPONSE:

Before Section 5 the last time the reader was introduced to the working
group model was on page 3, with the rather diffuse text, “The first call
for participation requested lightning talks, where each author could make a
brief statement about work that either had been done or was needed, with the
goal of contributing to the discussion of one or more working groups.” I
think what is lacking is some threading through of where these working
groups came from and how they are the focus of WSSSPE# going forward.

RESPONSE:

It is also distracting that each subsection of Section 5 represents a
subset/working group speaking with that voice, e.g., Section 5.3, “At
WSSSPE3 our group was interested in better understanding successful pathways
for scientific software to “transition” from grant-funded research
projects to industry sponsorship.” I think it makes clear sense for
individual appendices to speak with a “group” voice, but that the main
document should speak with a single authoritative voice.

RESPONSE:

To fix my perceived problem with Section 5, some background would need to be
added and some of the summary material about working groups that is
presently in the Conclusions section could be moved to introduce the working
groups better.

RESPONSE:

I just have one more comment on the narrative and analysis in this
summarizing manuscript. I suspect given the years WSSSPE has spent digesting
these topics and broadening and maturing its participants that it would be
worthwhile analyzing the following problem posed in the Conclusions a little
bit more in the current article: “We need to determine how to tie the
WSSSPE breakout activities to people’s jobs, so that they feel that
continuing them is a higher priority than it is now, perhaps through funding
the participants, or through funding coordinators for each activity, or
perhaps by getting the workshop participants to agree to a specific schedule
of activities during the workshop as we have tried to do in WSSSPE3.” Is
such a problem  symptomatic of all interdisciplinary projects with or
without limited funding? Would it be more successful if its efforts were
tied directly to the domain specific activities of the individual entities
represented by WSSSPE members? Is there a tension between the (idealistic?)
cross-domain goals of such a broad effort as WSSSPE and the practicalities
of getting people to work on someone else’s stuff?  I think there is an
opportunity for the authors to dig into their self assessed problem a little
bit more than they have.

RESPONSE:

Minor Comments:
1. I would avoid using the ableist term “bipolar” even if lifted
directly from the keynote;

RESPONSE: we've reworded this

2. I wonder if would be better to enhance the linking and referencing of
online materials in this document even beyond its already rich set of
footnoted links. For instance, all of the lightning talks refer to standing
proposals or active projects, but there are almost no reference URLs for
these projects or proposals;

RESPONSE:

3. There are many outlines in the appendices, and I’d like to note that
some outlines in the WSSSPE2 JORS article were negatively affected by poor
typesetting. You might want to make sure JORS gets them all right.

RESPONSE:

### 2016-08-26 WSSSPE3/report/WSSSPE3_review.tex
adding JORS reviews, a document for our responses, and making a couple of the easy changes

This report is based on collaborative notes taken during the workshop, which
were linked from the GitHub issues that represented the potential and actual
working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
lightning talks (\S\ref{sec:lightning}). The report also gives summaries of
action plans proposed by the working groups (\S\ref{sec:WGs}), and some
conclusions (\S\ref{sec:conclusions}). Lists of the organizing committee
(Appendix~\ref{sec:orgcom}), the registered attendees
(Appendix~\ref{sec:attendees}), and the travel award recipients
(Appendix~\ref{sec:awardees}) are compiled. Finally, the report includes longer
descriptions of the activities that occurred in each of the working groups that
made substantial progress
(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).

@@

WSSSPE3 left the traditional Supercomputing Conference environment this year,
and in Turk's words, this change spoke to the fact that scientific software
comes from many different types of inquiries, deployment, strategies for
maintenance, users, and ways of measuring the value of a piece of software. It
appeared to Turk that the supercomputing community generally adopts some
top-down approaches, whereas the SciPy community more often than not uses more
bottom-up systems. The essential messages perceived were also often bipolar: the
supercomputing community thinks that software is getting harder, with exascale
computing and optimization issues in mind; but the SciPy community thinks that
software is becoming better, with emerging tools such as Jupyter and
productivity packages for research workflows. Admitting such comparisons are
somewhat unfair generalizations, Turk reminded the audience that the different
approaches bring different types of ideas to the table, and he welcomed WSSSPE3
being conducted outside existing preconceptions.

>>>>

This report is based on collaborative notes taken during the workshop, which
were linked from the GitHub issues that represented the potential and actual
working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}), the keynote (\S\ref{sec:keynote}), and the lightning talks
presented at the meeting (\S\ref{sec:lightning}). The report also gives summaries of
action plans proposed by the working groups (\S\ref{sec:WGs}), and provides some
conclusions (\S\ref{sec:conclusions}). The first three appendices contain lists of the organizing committee
(Appendix~\ref{sec:orgcom}), the registered attendees
(Appendix~\ref{sec:attendees}), and the travel award recipients
(Appendix~\ref{sec:awardees}). Finally, the report includes longer
descriptions of the activities that occurred in each of the working groups that
made substantial progress
(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).

@@

WSSSPE3 left the traditional Supercomputing Conference environment this year,
and in Turk's words, this change spoke to the fact that scientific software
comes from many different types of inquiries, deployment, strategies for
maintenance, users, and ways of measuring the value of a piece of software. It
appeared to Turk that the supercomputing community generally adopts some
top-down approaches, whereas the SciPy community more often than not uses more
bottom-up systems. According to Turk, there is a divergence in views about
progress in software: the
supercomputing community thinks that software is getting harder, with exascale
computing and optimization issues in mind; but the SciPy community thinks that
software is becoming better, with emerging tools such as Jupyter and
productivity packages for research workflows. Admitting such comparisons are
somewhat unfair generalizations, Turk reminded the audience that the different
approaches bring different types of ideas to the table, and he welcomed WSSSPE3
being conducted outside existing preconceptions.

### 2016-08-26 WSSSPE3/report/wg_appendix_industry_interaction.tex
minor changes
adding JORS reviews, a document for our responses, and making a couple of the easy changes

\item A project could be \textbf{refunded}, and development or maintenance of
the software continue as planned.

@@

Next, we imagined \textbf{potential outcomes} of research on this topic, involving 
a set of case studies that look at successful and unsuccessful
transitions of researchers between academia and industry. This might address 
each of the transition types (to be described below). Successful transitions are
described as those that lead to either weak or strong sustainability (also
defined below). In addition, the results from this research might help create a 
generalizable framework that might allow for the study of different transition 
pathways (other than academia to industry).

@@

An example of data collection will be: 
\begin{itemize}
\item Origin: Where did project start? 
\item People involved: How many people in original project were involved in
transition/collaboration?
\item Specs on software
\item Language
\item Size
\item Hardness (age)
\item Lead-up to Transition: How long was project in development before it began transition?
\item Motivation for Transition: Why was transition initiated? By whom? 
\end{itemize}

>>>>

\item A project could be \textbf{re-funded}, and development or maintenance of
the software continue as planned.

@@

Next, we imagined \textbf{potential outcomes} of research on this topic, involving 
a set of case studies that look at successful and unsuccessful
transitions of researchers between academia and industry. This might address 
each of the transition types (as described below). Successful transitions are
described as those that lead to either weak or strong sustainability (also
defined below). In addition, the results from this research might help create a 
generalizable framework that might allow for the study of different transition 
pathways (other than academia to industry).

@@

An example of data collection is: 
\begin{itemize}
\item Origin: Where did project start? 
\item People involved: How many people in original project were involved in
transition/collaboration?
\item Specs on software
\item Language
\item Size
\item Hardness (age)
\item Lead-up to Transition: How long was project in development before it began transition?
\item Motivation for Transition: Why was transition initiated? By whom? 
\end{itemize}
